% !TeX root = main.tex
\documentclass[11pt]{article}

% Packages
\usepackage{amsmath, amssymb, amsfonts}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{physics}
\usepackage{mathtools}
\usepackage{lmodern}
\usepackage{firasans}
\usepackage{sectsty}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, calc}

% Custom design
\titlespacing*{\section}{0pt}{3.5ex plus 1ex minus .2ex}{2.3ex plus .2ex}
\fancyhf{}
\rhead{\thepage}
\lhead{\sffamily\nouppercase{\leftmark}}

% Geometry
\geometry{
    margin=1in
}
% \graphicspath{{figures/}}
\bibliographystyle{plain}


% Custom macros
\input{macros.tex}

% Title
\title{Gemma3, Architecture and Mathematical Foundations}
\author{Saman Keon}
% \date{\today}

\begin{document}

\maketitle
\pagestyle{fancy}
\thispagestyle{fancy}


% \begin{abstract}
% \input{sections/abstract.tex}
% \end{abstract}

\input{sections/introduction.tex}




\input{sections/architecture.tex}

\input{sections/layer.tex}
\input{sections/attention.tex}
\input{sections/rope.tex}
\input{sections/mlp.tex}
\input{sections/rmsnorm.tex}

\section{Conclusion}
The codebase accompanying this report is completely written by human.
Similarly I started writing this technical document all manually, without help of AI. But soon
learned that writing the LaTeX formulas can be daunting and error prone. Not only that, if I wanted
a diagram rich document, I needed to ask an AI to help; else diagrams would be mediocre at best.
So, sorry that this work is not purely human's work; though formulas and diagrams ideas came from me.

My next endeavor would be to implement Gemma with multi modal capabilities. I may come back and add some sections here, or publish a separate report. Will think about it.

\bibliography{references}

\end{document}
